\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-06 paper.tex
tweaks

% FIXME this is weak, but gives the basic outline
\begin{abstract}
New developments have made it possible to infer genetic genealogies in
the presence of recombination at scale, enabling many
downstream applications in population and statistical genetics.
The structure representing such recombinant genetic ancestry
is usually referred to as an ancestral recombination graph (ARG),
although there is some confusion about the interpretation and little
agreement on concrete details.
We propose a concrete definition of ARGs
in terms of genomes and their intervals of genetic inheritance (gARG)
and contrast this with the classical event-based definition (eARG).
The eARG definition is based on the stochatic process of the coalescent with recombination,
and we show how the assumptions of that process
place limitations on the patterns of inheritance that can be
represented, which are inadequate for modern large-scale datasets.
The eARG representation also requires complete precision about the ordering
of all recombination events, which we will rarely have sufficient
information to infer.
We show, in contrast, that the gARG definition
can fully capture the richness of modern large-scale datasets,
enables fine-grained
levels of precision about recombination to be represented,
and forms the basis of an efficient computational framework.
\end{abstract}

@@

This view of ARGs,
decoupled from generative models and
without the hard requirement
of complete precision on all historical events, may clarify inference goals and improve
methods for evaluation.
In most cases,
ARG inference is evaluated by simulating data from a known ground truth ARG, and comparing this to the inferred version via pairwise comparison of local trees along the genome, for example
using tree distance
metrics~\citep[e.g.][]{robinson1981comparison,kendall2016mapping},
as described by \citet{kuhner2015assessing}.
% While this provides valuable insights into the method's performance,
% it is not clear how well these tree distance metrics reflect
% performance on real data.
In comparing tree-by-tree along the genome, the effects of recombination
are only incorporated in an indirect manner through the correlations
between the local trees, rather than through directly taking into account
the persistence of nodes and edges across multiple trees.
Tree distance metrics usually have $O(n^2)$ time complexity and therefore cannot be
applied to the very large sample sizes currently of interest.
A recent trend has been to move away from such tree distance-based approaches and to examine more properties of the inferred ARGs.
\citet{brandt2021evaluation} focussed on the
distributions of pairwise MRCA times; \citet{deng2021distribution} derived an approximation to the distribution of the 
waiting distances between local trees; \citet{ignatieva2023distribution}
derived the distribution of the genomic span of an edge and that of a
clade of samples. In each case, simulation studies demonstrated 
substantial differences between these quantities in simulated and reconstructed ARGs,
that were not captured using tree-by-tree comparisons.
% Not sure this is the right place for this sentence
These evaluations have been based on
ground truth data from highly idealised simulations, some 
with small sample sizes,
and the effects of the richness of real data
(beyond simplistic error models)
on modern biobank scale or multi-generational pedigree datasets
are almost entirely unknown.
A suite of true ARG evaluation metrics, that take into account more of the
global topology and can be applied to large ARGs, would be a
valuable addition to the field.

@@

Figure \ref{hudson_vs_bigARG} illustrates the more complex state space
of the little ARG process, as well as the extra events which occur in the big ARG process.
Moreover, it depicts the rates of common ancestors and recombination events in each
interval of time of the realisations.
In order to evaluate these rates, and hence the sampling probability
(see e.g.\ \cite[Equation (3)]{mahmoudi2022bayesian}),
it is necessary to know the number of lineages and number of extant links
available for recombination in each time interval.
This may not be obvious.  example, in the gARG encoding depicted in Figure \ref{fig-ancestry-resolution}
it is clear that a recombination takes place between nodes \textsf{b} and
\textsf{d} as well as \textsf{e}, but the exact time of the event is ambiguous,
and thus so is the number of lineages during the time interval. This information cannot be recovered
from the gARG encoding 
Conventions can be introduced to resolve such ambiguities;
for the likelihood-based inference algorithms for the coalescent with recombination,
the two parent nodes are typically created at the time of the recombination event.
See the appendix of \cite{baumdicker2021efficient} for details of evaluating
coalescent with recombination likelihoods using this convention.
This is also the interpretation depicted in
Figure \ref{hudson_vs_bigARG}, but it means that the two edges above node \textsf{b}
in Figure \ref{fig-ancestry-resolution} should correspond to only one lineage,
along which all $m-1$ links are available for recombination.
The lineage then splits into two at the time of nodes \textsf{d} and \textsf{e}.
Nodes \textsf{k}, \textsf{l}, and \textsf{m} in Figure \ref{fig-ancestry-resolution}
demonstrate that the same issue can affect the number of links available for recombination:
without an external convention, the exact time at which the trapped ancestral material on
node \textsf{k} ceases to be available for an effective recombination in the little ARG process.

>>>>

% FIXME this is weak, but gives the basic outline
\begin{abstract}
New developments have made it possible to infer genetic genealogies in
the presence of recombination at scale, enabling many
downstream applications in population and statistical genetics.
The structure representing such recombinant genetic ancestry
is usually referred to as an ancestral recombination graph (ARG),
although there is some confusion about the interpretation and little
agreement on the concrete details.
We propose a concrete definition of ARGs
in terms of genomes and their intervals of genetic inheritance (gARG)
and contrast this with the classical event-based definition (eARG).
The eARG definition is based on the coalescent with recombination
stochastic process,
and we show how the assumptions of that process
place limitations on the patterns of inheritance that can be
represented, which are inadequate for modern large-scale datasets.
The eARG representation also requires complete precision about the ordering
of all recombination events, which we will rarely have sufficient
information to infer.
We show, in contrast, that the gARG definition
can fully capture the richness of modern large-scale datasets,
enables fine-grained
levels of precision about recombination to be represented,
and forms the basis of an efficient computational framework.
\end{abstract}

@@

This view of ARGs,
decoupled from generative models and
without the hard requirement
of complete precision on all historical events, may clarify inference goals and improve
methods for evaluation.
In most cases,
ARG inference is evaluated by simulating data from a known ground truth ARG, and comparing this to the inferred version via pairwise comparison of local trees along the genome, for example
using tree distance
metrics~\citep[e.g.][]{robinson1981comparison,kendall2016mapping},
as described by \citet{kuhner2015assessing}.
% While this provides valuable insights into the method's performance,
% it is not clear how well these tree distance metrics reflect
% performance on real data.
In comparing tree-by-tree along the genome, the effects of recombination
are only incorporated in an indirect manner through the correlations
between the local trees, rather than through directly taking into account
the persistence of nodes and edges across multiple trees.
Tree distance metrics usually have $O(n^2)$ time complexity and therefore cannot be
applied to the very large sample sizes currently of interest.
A recent trend has been to move away from such tree distance-based approaches and to examine more properties of the inferred ARGs.
\citet{brandt2021evaluation} focussed on the
distributions of pairwise MRCA times; \citet{deng2021distribution} derived an approximation to the distribution of the 
waiting distances between local trees; \citet{ignatieva2023distribution}
derived the distribution of the genomic span of an edge and that of a
clade of samples. In each case, simulation studies demonstrated 
substantial differences between these quantities in simulated and reconstructed ARGs,
that were not captured using tree-by-tree comparisons.
% Not sure this is the right place for this sentence
These evaluations have been based on
ground truth data from highly idealised simulations, with sample sizes limited to at most a few thousand (typically much fewer).
The effects of the richness of real data
on modern biobank-scale or multi-generational pedigree datasets
are almost entirely unknown (beyond studies using very simplistic error models).
A suite of ARG evaluation metrics, that take into account more of the
global topology and can be applied to large ARGs, would be a
valuable addition to the field.

@@

Figure \ref{hudson_vs_bigARG} illustrates the more complex state space
of the little ARG process, as well as the extra events which occur in the big ARG process.
Moreover, it depicts the rates of common ancestors and recombination events in each
interval of time of the realisations.
In order to evaluate these rates, and hence the sampling probability
(see e.g.\ \cite[Equation (3)]{mahmoudi2022bayesian}),
it is necessary to know the number of lineages and number of extant links
available for recombination in each time interval.
This may not be obvious. For example, in the gARG encoding depicted in Figure \ref{fig-ancestry-resolution}C
it is clear that a recombination takes place between nodes \textsf{b} and
\textsf{d} as well as \textsf{e}, but the exact time of the event is ambiguous,
and thus so is the number of lineages during the time interval. In fact, this
information cannot be recovered from the gARG encoding use in that figure,
so alternative conventions must be introduced to resolve such ambiguities;
for the likelihood-based inference algorithms for the coalescent with recombination,
the two parent nodes are typically created at the time of the recombination event.
See the appendix of \cite{baumdicker2021efficient} for details of evaluating
coalescent with recombination likelihoods using this convention.
This is also the interpretation depicted in
Figure \ref{hudson_vs_bigARG}, but it means that the two edges above node \textsf{b}
in Figure \ref{fig-ancestry-resolution} should correspond to only one lineage,
along which all $m-1$ links are available for recombination.
The lineage then splits into two at the time of nodes \textsf{d} and \textsf{e}.
Nodes \textsf{k}, \textsf{l}, and \textsf{m} in Figure \ref{fig-ancestry-resolution}
demonstrate that the same issue can affect the number of links available for recombination:
without an external convention, the exact time at which the trapped ancestral material on
node \textsf{k} ceases to be available for an effective recombination in the little ARG process.
